---
title: "Collate larval size data"
author: "Max Lindmark"
date: today
date-format: iso
toc: true
format: 
  html:
    page-layout: full
    embed-resources: true
knitr: 
  opts_chunk:
    fig.align: center
    out-width: 100%
editor: source
---

```{r load libraries}
#| message: false
#| warning: false

# Load libraries
library(tidyverse)
library(tidyterra)
library(tidync)
library(ggridges)
library(readxl)
library(janitor)
library(lubridate)
library(sdmTMB)
library(ncdf4)
library(patchwork)
library(terra)
library(viridis)
library(devtools)
library(ggsidekick)
theme_set(theme_sleek())
library(crayon)
library(marmap)
library(tidylog)

# Point to wd
home <- here::here()

# Load all custom functions in R/function
# - map-plot [source_url("https://raw.githubusercontent.com/maxlindmark/cod-interactions/main/R/functions/map-plot.R")]
# - callCopernicusCovariate
# - extractCovariateAtLocation
for (fun in list.files(paste0(home, "/R/functions"))) {
  source(paste(home, "R/functions", fun, sep = "/"))
}
```

# Explore data
## Read and clean data
### Old data

```{r}
#| message: false
#| warning: false

# 1992 -2008
length_old <- read_excel(paste0(home, "/data/larvea/1992_2010 MIK SWE Alla arter.xlsx"),
  sheet = 1,
  skip = 20
)

# Clean data!
length_old <- length_old |>
  mutate(species = fct_recode(species, "Chirolophis ascanii" = "Chirolophis  ascanii")) |>
  clean_names() |>
  pivot_longer(cols = x3:x130, names_to = "length_mm", values_to = "n") |>
  dplyr::select(year, day, month, haul, species, n, length_mm) |>
  mutate(length_mm = as.numeric(str_remove(length_mm, "x"))) |>
  # NA n means there was no recorded larvae in that size group (note there are also 0s in the data)
  drop_na(n) |>
  filter(n > 0) |>
  # expand the data so that 1 row is 1 individual
  # FIXME: the counts are not integers?! (hence the need for round)
  uncount(round(n)) |>
  dplyr::select(-n) |>
  # make a unique haul ID so that we can match with trawl data
  mutate(haul_id = paste(year, month, day, haul, sep = "_"))

# Read trawl data and match in coordinates
trawl_old <- read_excel(paste0(home, "/data/larvea/1992-2010 MIK SWE Tråldata.xlsx"),
  sheet = 1,
  skip = 8
) |>
  clean_names() |>
  # the last two coordinate columns are decimal degrees of haul position
  rename(
    haul = haul_no,
    lat = lat_decim_20,
    lon = long_decim_21
  ) |>
  # two rows without info, including year, so I'm dropping these
  drop_na(year) |>
  mutate(haul_id = paste(year, month, day, haul, sep = "_")) |>
  distinct(haul_id, .keep_all = TRUE) |>
  dplyr::select(haul_id, lat, lon, temp)

# Join trawl data to length data
length_old <- length_old |>
  left_join(trawl_old, by = "haul_id") |>
  dplyr::select(-haul) |>
  mutate(
    period = "old",
    day = as.numeric(day),
    month = as.numeric(month)
  )
```

### New data

```{r}
#| message: false
#| warning: false

# 2008-2022
length_new <- read_excel(paste0(home, "/data/larvea/ELDB(s) bara fisk 2008-2024.xlsx")) |>
  clean_names() |>
  mutate(
    species = fct_recode(species, "chirolophis ascanii" = "chirolophis  ascanii"),
    species = str_to_sentence(species)
  ) |>
  rename(length_mm = length) |>
  dplyr::select(haul_id, species, length_mm, number) |>
  drop_na(length_mm) |>
  # expand the data frame with "number"
  uncount(round(number))

#trawl_new <- read_excel(paste0(home, "/data/larvea/ELDB 2008-2024.xlsx")) |>
trawl_new <- read_excel(paste0(home, "/data/larvea/ELDB 2008-2024 - rättade djup.xlsx")) |>
  clean_names() |>
  rename(
    lat = start_latitud,
    lon = start_longitud
  ) |>
  dplyr::select(year, day, month, haul_id, lat, lon, sur_temp) |>
  rename(temp = sur_temp)

length_new <- length_new |>
  left_join(trawl_new, by = "haul_id") |>
  mutate(
    period = "new",
    year = as.numeric(year),
    day = as.numeric(day),
    month = as.numeric(month)
  )
```

### Join old and new

```{r}
#| message: false
#| warning: false
overlapping_yrs <- intersect(length_new$year, length_old$year)

length_old <- length_old |> filter(!year %in% c(overlapping_yrs))

d <- bind_rows(length_new, length_old) |>
  mutate(yday = yday(paste(year, month, day, sep = "-"))) |>
  filter(lon > 8) |>
  drop_na(lat) |>
  rename(temp_obs = temp)

# Add km UTM coords
d <- d |>
  add_utm_columns(ll_names = c("lon", "lat"))

sort(unique(d$species))

d <- d |> mutate(species = ifelse(species == "Ammodytes marinus", "Ammodytidae", species))
```

## Explore data

```{r}
#| message: false
#| warning: false
# Sample size
plot_map_fc +
  geom_point(
    data = d |>
      group_by(haul_id, Y, X, year) |>
      summarise(n = n()),
    aes(X * 1000, Y * 1000, color = n),
    size = 0.5
  ) +
  facet_wrap(~year, ncol = 8) +
  scale_color_viridis(trans = "sqrt") +
  ggtitle("Sample size per haul")

# Day of the year
plot_map_fc +
  geom_point(
    data = d |>
      distinct(haul_id, .keep_all = TRUE),
    aes(X * 1000, Y * 1000, color = yday),
    size = 0.5
  ) +
  facet_wrap(~year, ncol = 8) +
  scale_color_viridis(trans = "sqrt") +
  ggtitle("Day of the year of sampling in space")

# Which dates are sampled?
d |>
  ggplot(aes(as.factor(month))) +
  scale_fill_viridis(discrete = TRUE) +
  geom_histogram(stat = "count")

d |>
  ggplot(aes(x = yday, y = as.factor(year), fill = after_stat(x))) +
  scale_fill_viridis(alpha = 0.8, name = "") +
  geom_density_ridges_gradient(alpha = 0.75) +
  theme_facet_map() +
  labs(y = "year")

# Species
sort(unique(d$species))

# Clean species names!
d <- d |>
  mutate(species = str_to_sentence(species))

sort(unique(d$species))

d |>
  group_by(year, species) |>
  summarise(n = n()) |>
  ggplot(aes(year, n, fill = species)) +
  guides(fill = "none") +
  facet_wrap(~species, scales = "free_y") +
  geom_bar(stat = "identity")

# Filter species with at least 6 years of data and minimum 5 sizes per year and 10 samples
d <- d |>
  # At least 5 samples by year
  mutate(n = n(), .by = c(species, year)) |>
  filter(n >= 10) |>
  mutate(n_years = length(unique(year)), .by = species) |>
  filter(n_years >= 7) |>
  dplyr::select(-n_years, -n)

sort(unique(d$species))
```

## Trim data

Sizes are here transition to post-larvae (after which they metamorphose into juveniles). In some cases, the distribution of sizes overlap a lot with the post-larve size. In those cases we go for the size where they move from pelagic to benthic habitats (which often is what splits the size-distributions). Some species are also not larvae.

```{r}
sort(unique(d$species))

# Agonus cataphractus: Trim the 3 data points above postlarvale sizes of 20 mm
d <- d |> filter(!(species == "Agonus cataphractus" & length_mm > 20))

# Ammodytidae: Two species, don't foget to add ammodyteas into this group in the data processing script, since identification to species level likely has changed over time! There also appears to be two cohorts. Post-larvae cutoff at 26-30 mm cutoff at post-larvae for the two species, but here perhaps we want to split around 55 mm to separate the clusters?

d |>
  filter((species == "Ammodytidae" & length_mm < 57)) |>
  ggplot(aes(length_mm)) +
  geom_histogram()

d <- d |> filter(!(species == "Ammodytidae" & length_mm > 58))

# Anguilla anguilla: Ok
# Aphia minuta: not larvae but can include anyway. Remove the one outlier
d |>
  filter(species == "Aphia minuta") |>
  ggplot(aes(length_mm)) +
  geom_histogram()

d <- d |> filter(!(species == "Aphia minuta" & length_mm > 60))

# Argentina silus: Merge with Argentina spp? Seems like identification to species level changed over time. Post larvae under 50 which seems fitting for a cutoff size
d |>
  filter(species %in% c("Argentina silus", "Argentina spp")) |>
  ggplot(aes(length_mm, fill = species)) +
  geom_histogram()

d <- d |>
  filter(!(species %in% c("Argentina silus", "Argentina spp") & length_mm > 50)) |>
  mutate(species = ifelse(species == "Argentina silus", "Argentina spp", species))

# Chirolophis ascanii: Post-larvae around 20 mm. However, data indicate a size for metamorphosis around 30 mm (since they then settle on the seabead).
# d |>
#   filter(species == "Chirolophis ascanii") |>
#   ggplot(aes(length_mm)) +
#   geom_histogram()
#
# d <- d |> filter(!(species == "Chirolophis ascanii" & length_mm > 30))

# Clupea harengus: Post-larve until 48-50 mm, seems fitting
d |>
  filter(species %in% c("Clupea harengus")) |>
  ggplot(aes(length_mm, fill = species)) +
  geom_histogram()

d <- d |> filter(!(species == "Clupea harengus" & length_mm > 50))

# Crystallogobius linearis: Adult, otherwise seems OK. Malin was going to check something with the gear: are we sampling a different part of the size-distribution now?
# Enchelyopus cimbrius: Potentially different cohorts because spawning time is long and potentially not accurate in the literature given the sizes we observe here. <span style="color:red;">Drop this from the study</span>
d <- d |> filter(!(species == "Enchelyopus cimbrius"))

# Limanda limanda: Post-larve may remain pelagic until 30 mm, use that as a cutoff.
d |>
  filter(species %in% c("Limanda limanda")) |>
  ggplot(aes(length_mm, fill = species)) +
  geom_histogram()

d <- d |> filter(!(species == "Limanda limanda" & length_mm > 30))

# Maurolicus muelleri: These are not larvae and not spawning frequently here. <span style="color:red;">Drop this from the study</span>
d <- d |> filter(!(species == "Maurolicus muelleri"))

# Microstomus kitt: metamorphosis at 18 mm, but still pelagic? we judge that they are pelagic until 40 mm
d |>
  filter(species %in% c("Microstomus kitt")) |>
  ggplot(aes(length_mm, fill = species)) +
  geom_histogram()

d <- d |> filter(!(species == "Microstomus kitt" & length_mm > 40))

# Myoxocephalus scorpius: Pelagic until about 20 mm, use that as a cutoff
d |>
  filter(species %in% c("Myoxocephalus scorpius")) |>
  ggplot(aes(length_mm, fill = species)) +
  geom_histogram()

d <- d |> filter(!(species == "Myoxocephalus scorpius" & length_mm > 20))

# Pholis gunnellus: pelagic until 35 mm, use that as a cutoff
d |>
  filter(species %in% c("Pholis gunnellus")) |>
  ggplot(aes(length_mm, fill = species)) +
  geom_histogram()

d <- d |> filter(!(species == "Pholis gunnellus" & length_mm > 35))

# Pomatoschistus sp: Not larve, could keep for the same reason as other non-larvae species

# Sardina pilchardus: Ok
# Sprattus sprattus: Too many adults because they don't spawn at the right time. If we trim to postlarvae, then we have too few anyway. <span style="color:red;">Drop this from the study</span>

d <- d |> filter(!species == "Sprattus sprattus")

d <- d |> filter(!species == "Sprattus sprattus adult")

# Syngnathus rostellatus: No larval stage, can keep for the same reason as other non-larvae

# Taurulus bubalis: 12 mm cutoff (is that post-larva maximum size or start?)
d |>
  filter(species %in% c("Taurulus bubalis")) |>
  ggplot(aes(length_mm, fill = species)) +
  geom_histogram()

d <- d |> filter(!(species == "Taurulus bubalis" & length_mm > 35))

# Add in which species are not strictly larvae
d <- d |>
  mutate(life_stage = ifelse(species %in% c(
    "Ammodytidae", "Syngnathus rostellatus",
    "Pomatoschistus sp", "Crystallogobius linearis",
    "Aphia minuta"
  ),
  "mixed", "larvae"
  ))

# Plot species samples in space, color by year
plot_map_fc +
  geom_point(
    data = d,
    aes(X * 1000, Y * 1000, color = year),
    size = 0.5
  ) +
  facet_wrap(~species, ncol = 6) +
  scale_color_viridis() +
  ggtitle("Samples in space by species and year")

d |>
  group_by(year, species) |>
  summarise(n = n()) |>
  ggplot(aes(year, n, fill = species)) +
  guides(fill = "none") +
  facet_wrap(~species, scales = "free_y", ncol = 5) +
  geom_bar(stat = "identity") +
  scale_fill_viridis(discrete = TRUE)

d |>
  distinct(species, period) |>
  as.data.frame()
```

## Add covariate to hauls

```{r}
# Specify covariates path for simplicity
covPath <- paste0(home, "/data/covariates")
```

### Satellite derived temperatures
https://data.marine.copernicus.eu/product/SST_BAL_SST_L4_REP_OBSERVATIONS_010_016/description

```{r}
## Load satellite derived SST.
# Source: https://data.marine.copernicus.eu/product/SST_BAL_SST_L4_REP_OBSERVATIONS_010_016/download
# Print details
print(nc_open(paste(covPath, "sst", "DMI_BAL_SST_L4_REP_OBSERVATIONS_010_016_1711802008633.nc", sep = "/")))

# Load and gather the temperature data in a tibble
temp_tibble <- callCopernicusCovariate("sst", messages = 1)
```

```{r}
# Visualize temperature frequency distribution
hist(temp_tibble$sst)

# Visualize temperature spatial distribution
# plot_map +
#   geom_point(data = temp_tibble,
#              aes(X*1000, Y*1000, color = sst))

# Obtain temporal availability, this will be the temporal window to filter the data
unique(temp_tibble$year)

# Trim years we have temperature for (again, annoying! Fix the temperatures later)
d <- d |>
  filter(year %in% unique(temp_tibble$year))

# Loop through all year combos, extract the temperatures at the data locations
d <- extractCovariateAtLocation(
  "sst", # Name of the covariate to extract. One of: sst, chlorophyll, depth.
  d, # A df containing the set of yearand locations to be evaluated.
  temp_tibble, # A df containing the covariate at location
  changesYearly = 1, # Is the covariate time variant (e.g. temp) or not (e.g. depth)
  "temp", # Name to give to the covariate evaluated at location in the df
  messages = 1 # dichotomous
)
```

### Satellite derived chlorophyll abundance
https://data.marine.copernicus.eu/product/GLOBAL_MULTIYEAR_BGC_001_029/description

```{r}
## Load satellite derived chlorophyll
# Source: https://data.marine.copernicus.eu/product/GLOBAL_MULTIYEAR_BGC_001_029/download
# Print details
print(nc_open(paste(covPath, "chlorophyll", "cmems_mod_glo_bgc_my_0.25_P1D-m_1713795613611_01012017_12312022.nc", sep = "/")))

# Load and gather the temperature data in a tibble
chl_tibble <- callCopernicusCovariate("chlorophyll", messages = 1)
```

```{r}
# Visualize chlorophyll frequency distribution
hist(chl_tibble$chl)

# Visualize chlorophyll spatial distribution
# plot_map +
#   geom_point(data = chl_tibble,
#              aes(X*1000, Y*1000, color = chl))

# Obtain temporal availability, this will be the temporal window to filter the data
sort(unique(chl_tibble$year))

# Trim years we have chlorophyll for.
d <- d |>
  filter(year %in% unique(chl_tibble$year)) # We loose 13% of the data by including chl.

# Loop through all year combos, extract the chl at the data locations
d <- extractCovariateAtLocation(
  "chl",
  d,
  chl_tibble,
  changesYearly = 1,
  "chl",
  messages = 1
)
```

### Satellite derived depth
https://data.marine.copernicus.eu/product/GLOBAL_MULTIYEAR_BGC_001_029/description

```{r}
# Generate a depth box containing the bathymetries.
depth_box <- getNOAA.bathy(min(d$lon) - .1, max(d$lon) + .1, min(d$lat) - .1, max(d$lat) + .1)

# Obtain depth at locations.
d <- cbind(
  d,
  get.depth(depth_box, x = d$lon, y = d$lat, locator = F)["depth"]
)

## Convert to strictly positive values.
d$depth <- d$depth * (-1)

# Check
plot_map +
  geom_point(
    data = d,
    aes(X * 1000, Y * 1000, color = depth)
  )

d <- d |> tidylog::filter(depth > 0)
```

## Check covariates 

```{r}
# Get the proportion of observations not assigned with a covariate value at prior steps
colMeans(is.na(d))
```

## Plot response variables

```{r}
#| message: false
#| warning: false
d |>
  summarise(n = n(), .by = species) |>
  arrange(desc(n))

# Distribution of data
ggplot(d, aes(length_mm)) +
  geom_histogram() +
  facet_wrap(~species, scales = "free")

# Effect of day of the year
ggplot(d, aes(yday, length_mm)) +
  geom_point(size = 0.4, alpha = 0.4) +
  geom_smooth(method = "lm") +
  facet_wrap(~species, scales = "free")

# Effect of year
ggplot(d, aes(year, length_mm)) +
  geom_point(size = 0.4, alpha = 0.4) +
  geom_smooth(method = "lm") +
  facet_wrap(~species, scales = "free")

# Effect of temperature
ggplot(d, aes(temp, length_mm)) +
  geom_point(size = 0.4, alpha = 0.4) +
  geom_smooth(method = "lm") +
  # geom_smooth() +
  facet_wrap(~species, scales = "free")

# Effect of chlorophyll
ggplot(d, aes(chl, length_mm)) +
  geom_point(size = 0.4, alpha = 0.4) +
  geom_smooth(method = "lm") +
  # geom_smooth() +
  facet_wrap(~species, scales = "free")
```

## Save data

```{r}
d <- d |> drop_na(temp)

write_csv(d, paste0(home, "/data/clean/larval_size.csv"))
```
